Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed-particle hydrodynamics

### 2017-02-24 Manuscripts_GMD.tex
revision according to Dr. Bursik's comments are made

%%xx General points
%%xx 0 Some suggestions in the text itself are shown with "xx" -- so if you just search for "xx" you will find them --> Done
%%xx 1 Make sure tenses match throughout
%%xx 2 use \citet for citations in-line (text), e.g., "\citet{monaghanpaper} found that SPH is great...", not "Monaghan \citep{monaghanpaper} found that SPH is great..." -->Done
%%xx 3 Need a section on the atmospheric structure, and how atmosphere can be incorporated into model, especially NWP or radiosondes
%%xx 4 You often seem to forget to put a space between paragraphs.  LaTeX of course does not recognize a <return> as a paragraph delimiter, nor is it proper to end a normal paragraph with \\ or % -->Done
%%xx 5 You are inconsistent with capitalization in section headings.  Either always use an initial capital letter, or only use an initial capital letter on the first word, proper names or acronyms -->Done
%%xx 6 Figures are of generally poor quality -- use PDF and reproduce at larger scale.  
%%xx 7 Difficult to understand what figure of Pinatubo plume is supposed to be telling us about the model

@@

\introduction  %% \introduction[modified heading if necessary]
%<<<<<<< HEAD
\subsection{Volcanic ash hazards}
Primary hazards associated with explosive volcanic eruptions include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats to aviation posed by volcanic ash in the atmosphere
Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our focus here is the hazards that volcanic ash poses to aircraft and communities.

@@

\subsection{Existing plume models}
Several 1D (one dimensional) volcanic plume models have been developed in the past few decades. Ranging from the most basic 1D model \citep{woods1988fluid} which only accounts for mass conservation to more recently developed 1D models (eg. \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity}) which tend to account for more complex physics effects. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, in these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is also ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models require little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcanic plume development (such as \citep{267, 1194, 3541}), combined with VATDs (such as \citep{114, draxler2015hysplit}) are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skill in hazards forecasts especially where the plume model is used to generate source conditions for complex VATD models requires 2D (two dimensional) and 3D models.

@@

While the SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM ( Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation in 2006 \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still pretty coarse compared with SK-3D and PDAC.

@@

To summarize, each 3D model has its own focus based on problem of interest and modeling/numerical choices made.
 %Different simplification was made based on their scope. Improvements have been making since their birth. 
 Accuracy of simulation and forecast skill (depends on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depends on number of governing equations, resolution, numerical methods and parallel techniques).%are always a couple of conflict considerations in 3D plume simulations.\\

\subsection{Features of SPH}
All of these existing 3D models are mesh based Eulerian methods. To date there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed-particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
and,
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These  will be enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcanic plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})). So advection term will be treated exactly with SPH \citep{monaghan2005smoothed}.
\item It is easy to include various physics effect (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases will lead to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

@@

As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and physics heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases will not show up under this assumption). Immediate dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term will not show up with this assumption).

@@

\subsection{Boundary conditions}
In the current model the initial domain is a box. The boundaries are categorized as the velocity inlet (a circular area at the center of the bottom of the box), wall boundary (box bottom), pressure outlet (other faces of the box).

@@

\subsection{Fundamental principles}
There are several procedures for discretizing governing equations (PDEs or ODEs) with SPH. There is slight difference between discretized formalism obtained by different routes. We present here one of them following \citet{monaghan1992smoothed, monaghan2005smoothed, monaghan2012smoothed}. The starting point of approximating a function with SPH is the translation property of the Dirac function.
\begin{equation}
A(\textbf{x})=\int_{-\infty}^{\infty} A(\textbf{x} \prime) \delta (\textbf{x} \prime - \textbf{x}) d\textbf{x} \prime
\label{eq:Dirac-translation}
\end{equation}
The Dirac function in Eq. (\ref{eq:Dirac-translation}) can be approximated by a weighting function $w(\textbf{x}-\textbf{x}\prime, h)$ (or $w(\textbf{x}\prime-\textbf{x}, h)$) which will tend to a Dirac function when $h \rightarrow 0$ :
\begin{equation}
\lim _{h \rightarrow 0} w(\textbf{x} \prime-\textbf{x}, h) =  \delta (\textbf{x} \prime - \textbf{x})
\label{eq:SPH_kernel_delta}
\end{equation}
The function $A(\textbf{x})$ then can be approximated by:
\begin{equation}
A(\textbf{x}) \approx <A(\textbf{x})> = \int_{\Omega} A(\textbf{x} \prime) w(\textbf{x}-\textbf{x}\prime, h) d\textbf{x}\prime + O(h^2)
\label{eq:SPH-fundamental-principle}
\end{equation}
Where $h$ is the smoothing length, determining the interaction distance. According to Eq. (\ref{eq:SPH-fundamental-principle}), the order of accuracy of SPH method is second order. However, in practice, second order can not be achieved because there is no guarantee on the symmetry of particle distribution in real simulation \citep{price2012smoothed}.
Recall that $d\textbf{x}\prime = \dfrac{dm \prime}{\rho \prime}$, the integration equation, Eq. (\ref{eq:SPH-fundamental-principle}), can be approximated by summation and lead to an approximation of the function $A$:
\begin{equation}
<A(\textbf{x})> \approx \sum_b m_b \dfrac{A_b}{\rho_b} w(\textbf{x}-\textbf{x}_b, h)
\label{eq:SPH-approximation-sum}
\end{equation}
where the summation is over all the particles within the region of compact support of the weighting function. 
Gradient terms may be straightforwardly calculated by taking the derivative of Eq. (\ref{eq:SPH-approximation-sum}), giving
\begin{equation}
\begin{split}
<\nabla A(\textbf{x})> & = \dfrac{\partial }{\partial \textbf{x}} \int_{\Omega} A(\textbf{x} \prime) w(\textbf{x}-\textbf{x}\prime, h) d\textbf{x}\prime + O(h^2) \\
& \approx \sum_b m_b \dfrac{A_b}{\rho_b} \nabla w(\textbf{x} - \textbf{x}_b, h)
\end{split} 
\label{eq:SPH-scalar-function-gradient}
\end{equation}
For vector quantities the expressions are similar, simply replacing $A$ with $\textbf{A}$ in Eq. (\ref{eq:SPH-approximation-sum}) and and Eq. (\ref{eq:SPH-scalar-function-gradient}), giving
\begin{align}
<\textbf{A}(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} w(\textbf{x}-\textbf{x}_b, h) \\
<\nabla \cdot \textbf{A}(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \cdot \nabla w(\textbf{x} - \textbf{x}_b, h) \\
<\nabla \times \textbf{A}(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \times \nabla w(\textbf{x} - \textbf{x}_b, h) \\
<\nabla^j \textbf{A}^i(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b^i}{\rho_b} \nabla^j w(\textbf{x} - \textbf{x}_b, h) 
\label{eq:SPH-vecctor-function}
\end{align}
%

@@

\subsection{Discretization of governing equations and extensibility}
The basic interpolation given in Eq. (\ref{eq:SPH-approximation-sum}) to Eq. (\ref{eq:SPH-vecctor-function}) provides a general way to obtaining SPH expressions of governing equations. The problem is that using these expressions ``{as is}" in general leads to quite poor gradient estimates. Various tricks are used to conserve linear and angular momentum and thermal energy \citep{monaghan1992smoothed}. Special treatments are also need for second order derivative terms. We only refer here to one of these possible discretizations of compressible Euler equations with SPH:
\begin{align}
<\rho_a> = \sum_b m_b w_{ab} (h) \label{eq:ns-sph-d} \\
<\dfrac{d \textbf{v}_a}{d t}>= \sum_b m_b (\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}) \nabla_a w_{a b}(h) +\textbf{g} \label{eq:ns-sph-v} \\
<\dfrac{d e_a}{d t}>=
 0.5\sum_b m_b \textbf{v}_{a b}(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}) \cdot \nabla_a w_{a b}(h) \label{eq:ns-sph-e}
\end{align}
where, $a$ is the SPH particle index, $\textbf{v}_{a b} = \textbf{v}_a - \textbf{v}_b$. $\Pi$ is an artificial viscosity term, which will be discussed in section \ref{sec:artificial-viscosity}.
However, these discretized formulations will not guarantee conservation of entropy. We can obtain conservation of entropy by giving up on conservation of mass or thermal energy adopting  alternative discretized formulations \citep{monaghan1992smoothed}. 
As a Lagrangian method, particle position will also be updated at every time step.
\begin{equation}
<\dfrac{d \textbf{x}_a}{dt}> = \textbf{v}_a \label{eq:SPH-update-pos}
\end{equation}

We highlight an important feature of the SPH methodology. Adding new physics and new phases into the model is trivial in terms of discretization. For example, adding of new source (or sink) into Eq.  (\ref{eq:ns-sph-d}), adding a drag force into Eq. (\ref{eq:ns-sph-v})  anding a heat exchange term into Eq. (\ref{eq:ns-sph-e}) will lead to the new discretization form:
\begin{align}
<\rho_a> = \sum m_b w_{ab} (h) + \dot{\rho}(\textbf{x},t)\label{eq:ns-source-sph-d} \\
<\dfrac{d \textbf{v}_a}{d t}>= \sum_b m_b (\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}) \nabla_a w_{a b}(h) +\textbf{g} + D \sum	_b m_b \dfrac{\textbf{v}_b - \textbf{v}_a}{\rho_b} \label{eq:ns-drag-sph-v} \\
<\dfrac{d e_a}{d t}>=
 0.5\sum_b m_b \textbf{v}_{a b}(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}) \cdot \nabla_a w_{a b}(h) + \sum_b \dfrac{m_b}{\rho_b}(\kappa_a + \kappa_b) \dfrac{(T_a - T_b)}{\textbf{r}_a - \textbf{r}_b} w_{ab}(h) \label{eq:ns-conduction-sph-e}
\end{align}
where the source term $\dot{\rho}$ can be a "sink" of erupted vapor due to phase change like in ATHAM \citep{oberhuber1998volcanic}. $D$ is a drag force coefficient. $\kappa$ is heat conduction coefficient. Other physics can be added easily in a similar way. Adding of these new terms will lead to modification of several lines in the source code. The drag force term should show up only when dynamic dis-equilibrium between different phases is considered. In that case, each phase will need one set of governing equations of Navier-Stokes type. Adding of new phases into SPH code only needs few new lines for the new phase besides adding  interaction terms.

@@

\subsection{Time step}
The physical properties (velocity, density, position and pressure) change every time step due to particle interactions. We have to consider the Courant condition, which is in spirit similar to the Courant condition for the grid-based methods, to determine the amount of the time step $\Delta t$.
\begin{equation}
\Delta t = CFL \min_a \bigg \lbrace \dfrac{[\frac{m_a}{\rho_a}]^{\frac{1}{d}}}{c_a} \bigg \rbrace
\end{equation}
Where $c_a$ is sound speed at particle a, $d$ is number of dimensions. First order Euler time integration, with $CFL = 0.2$ has been used in three dimensional calculations.

@@

Experiments have been carried out on CCR (the computational cluster of Center for Computational Research) at Buffalo. E5645 CPUs running at 2.40GHz clock rate with 4GB memory per core on a Q-Logic Infiniband is used in these tests. Each node comprise of two sockets with 6 of these cores. Memory and level 3 cache are shared on each node. The initial domain is $[-4.8km,4.8km] \times [-4.8km,4.8km] \times [0km, 6km]$. Almost linear speed up is observed in our strong scalability test (Fig. \ref{fig:strong_scale}).

>>>>

%%xx General points
%%xx 0 Some suggestions in the text itself are shown with "xx" -- so if you just search for "xx" you will find them --> Done
%%xx 1 Make sure tenses match throughout
%%xx 2 use \citet for citations in-line (text), e.g., "\citet{monaghanpaper} found that SPH is great...", not "Monaghan \citep{monaghanpaper} found that SPH is great..." -->Done
%%xx 3 Need a section on the atmospheric structure, and how atmosphere can be incorporated into model, especially NWP or radiosondes
%%xx 4 You often seem to forget to put a space between paragraphs.  LaTeX of course does not recognize a <return> as a paragraph delimiter, nor is it proper to end a normal paragraph with \\ or % -->Done
%%xx 5 You are inconsistent with capitalization in section headings.  Either always use an initial capital letter, or only use an initial capital letter on the first word, proper names or acronyms
%%xx 6 Figures are of generally poor quality -- use PDF and reproduce at larger scale.  
%%xx 7 Difficult to understand what figure of Pinatubo plume is supposed to be telling us about the model

@@

\introduction  %% \introduction[modified heading if necessary]
%<<<<<<< HEAD
\subsection{Volcanic ash hazards}
Primary hazards associated with explosive volcanic eruptions include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats to aviation posed by volcanic ash in the atmosphere.
Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our focus here is the hazards that volcanic ash poses to aircraft and communities.

@@

\subsection{Existing plume models}
Several 1D (one dimensional) volcanic plume models have been developed in the past few decades. Ranging from the most basic 1D model \citep{woods1988fluid} which only accounts for mass conservation to more recently developed 1D models  \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity} which tend to account for more complex physics effects. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, in these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is also ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models require little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcanic plume development \citep[such as][]{267, 1194, 3541}, combined with VATDs \citep[such as][]{114, draxler2015hysplit} are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skill in hazards forecasts especially where the plume model is used to generate source conditions for complex VATD models requires 2D (two dimensional) and 3D models.

@@

While the SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM ( Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still coarse compared with SK-3D and PDAC.

@@

To summarize, each 3D model has its own focus based on problem of interest and modeling/numerical choices made.
 %Different simplification was made based on their scope. Improvements have been making since their birth. 
 Accuracy of simulation and forecast skill (depends on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depends on number of governing equations, resolution, numerical methods and parallel techniques) are always a pair of conflict considerations in 3D plume simulations.\\

\subsection{Features of SPH}
All of these existing 3D models are mesh based Eulerian methods. To date there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed-particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item Better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These  will be enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcanic plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})). So advection term will be treated exactly with SPH \citep{monaghan2005smoothed}.
\item It is easy to include various physics effect (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases will lead to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

@@

As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases will not show up under this assumption). Immediate dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term will not show up with this assumption).

@@

\subsection{Boundary conditions}
In the current model the initial domain is a box. The boundaries are categorized as the velocity inlet (a circular area at the center of the bottom of the box), wall boundary (box bottom), pressure outlet (other faces of the box).
%probably should add a section to list the input parameters in the verification section

@@

\subsection{Fundamental principles}
There are several procedures for discretizing governing equations (PDEs or ODEs) with SPH. There is slight difference between discretized formalism obtained by different routes. We present here one of them following \citet{monaghan1992smoothed, monaghan2005smoothed, monaghan2012smoothed}. The starting point of approximating a function with SPH is the translation property of the Dirac function.
\begin{equation}
A(\textbf{x})=\int_{-\infty}^{\infty} A(\textbf{x} \prime) \delta (\textbf{x} \prime - \textbf{x}) d\textbf{x} \prime
\label{eq:Dirac-translation}
\end{equation}
The Dirac function in Eq. (\ref{eq:Dirac-translation}) can be approximated by a weighting function $w(\textbf{x}-\textbf{x}\prime, h)$ (or $w(\textbf{x}\prime-\textbf{x}, h)$) which will tend to a Dirac function when $h \rightarrow 0$ :
\begin{equation}
\lim _{h \rightarrow 0} w(\textbf{x} \prime-\textbf{x}, h) =  \delta (\textbf{x} \prime - \textbf{x})
\label{eq:SPH_kernel_delta}
\end{equation}
The function $A(\textbf{x})$ then can be approximated by:
\begin{equation}
A(\textbf{x}) \approx <A(\textbf{x})> = \int_{\Omega} A(\textbf{x} \prime) w(\textbf{x}-\textbf{x}\prime, h) d\textbf{x}\prime + O(h^2)
\label{eq:SPH-fundamental-principle}
\end{equation}
Where $h$ is the smoothing length, determining the interaction distance. According to Eq. (\ref{eq:SPH-fundamental-principle}), the order of accuracy of SPH method is second order. However, in practice, second order can not be achieved because there is no guarantee on the symmetry of particle distribution in real simulation \citep{price2012smoothed}.
Recall that $d\textbf{x}\prime = \dfrac{dm \prime}{\rho \prime}$, the integration equation, Eq. (\ref{eq:SPH-fundamental-principle}), can be approximated by summation and lead to an approximation of the function $A$:
\begin{equation}
<A(\textbf{x})> \approx \sum_b m_b \dfrac{A_b}{\rho_b} w(\textbf{x}-\textbf{x}_b, h)
\label{eq:SPH-approximation-sum}
\end{equation}
where the summation is over all the particles within the region of compact support of the weighting function. 
Gradient terms may be straightforwardly calculated by taking the derivative of Eq. (\ref{eq:SPH-approximation-sum}), giving
\begin{equation}
\begin{split}
<\nabla A(\textbf{x})> & = \dfrac{\partial }{\partial \textbf{x}} \int_{\Omega} A(\textbf{x} \prime) w(\textbf{x}-\textbf{x}\prime, h) d\textbf{x}\prime + O(h^2) \\
& \approx \sum_b m_b \dfrac{A_b}{\rho_b} \nabla w(\textbf{x} - \textbf{x}_b, h)
\end{split} 
\label{eq:SPH-scalar-function-gradient}
\end{equation}
For vector quantities the expressions are similar, simply replacing $A$ with $\textbf{A}$ in Eq. (\ref{eq:SPH-approximation-sum}) and Eq. (\ref{eq:SPH-scalar-function-gradient}), giving
\begin{align}
<\textbf{A}(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} w(\textbf{x}-\textbf{x}_b, h) \\
<\nabla \cdot \textbf{A}(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \cdot \nabla w(\textbf{x} - \textbf{x}_b, h) \\
<\nabla \times \textbf{A}(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \times \nabla w(\textbf{x} - \textbf{x}_b, h) \\
<\nabla^j \textbf{A}^i(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b^i}{\rho_b} \nabla^j w(\textbf{x} - \textbf{x}_b, h) 
\label{eq:SPH-vecctor-function}
\end{align}
%

@@

\subsection{Discretization of governing equations and extensibility}
The basic interpolation given in Eq. (\ref{eq:SPH-approximation-sum}) to Eq. (\ref{eq:SPH-vecctor-function}) provides a general way to obtaining SPH expressions of governing equations. The problem is that using these expressions ``{as is}" in general leads to quite poor gradient estimates. Various tricks are used to conserve linear and angular momentum and thermal energy \citep{monaghan1992smoothed}. Special treatments are also needed for second order derivative terms. We only refer here to one of these possible discretizations of compressible Euler equations with SPH:
\begin{align}
<\rho_a> = \sum_b m_b w_{ab} (h) \label{eq:ns-sph-d} \\
<\dfrac{d \textbf{v}_a}{d t}>= \sum_b m_b (\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}) \nabla_a w_{a b}(h) +\textbf{g} \label{eq:ns-sph-v} \\
<\dfrac{d e_a}{d t}>=
 0.5\sum_b m_b \textbf{v}_{a b}(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}) \cdot \nabla_a w_{a b}(h) \label{eq:ns-sph-e}
\end{align}
where, $a$ is the SPH particle index, $\textbf{v}_{a b} = \textbf{v}_a - \textbf{v}_b$. $\Pi$ is an artificial viscosity term, which will be discussed in section \ref{sec:artificial-viscosity}.
However, these discretized formulations will not guarantee conservation of entropy. We can obtain conservation of entropy by giving up on conservation of mass or thermal energy adopting  alternative discretized formulations \citep{monaghan1992smoothed}. 
As a Lagrangian method, particle position will also be updated at every time step.
\begin{equation}
<\dfrac{d \textbf{x}_a}{dt}> = \textbf{v}_a \label{eq:SPH-update-pos}
\end{equation}

We highlight an important feature of the SPH methodology. Adding new physics and new phases into the model is trivial in terms of discretization. For example, adding of new source (or sink) into Eq.  (\ref{eq:ns-sph-d}), adding a drag force into Eq. (\ref{eq:ns-sph-v})  and adding a heat exchange term into Eq. (\ref{eq:ns-sph-e}) will lead to the new discretization form:
\begin{align}
<\rho_a> = \sum m_b w_{ab} (h) + \dot{\rho}(\textbf{x},t)\label{eq:ns-source-sph-d} \\
<\dfrac{d \textbf{v}_a}{d t}>= \sum_b m_b (\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}) \nabla_a w_{a b}(h) +\textbf{g} + D \sum	_b m_b \dfrac{\textbf{v}_b - \textbf{v}_a}{\rho_b} \label{eq:ns-drag-sph-v} \\
<\dfrac{d e_a}{d t}>=
 0.5\sum_b m_b \textbf{v}_{a b}(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}) \cdot \nabla_a w_{a b}(h) + \sum_b \dfrac{m_b}{\rho_b}(\kappa_a + \kappa_b) \dfrac{(T_a - T_b)}{\textbf{r}_a - \textbf{r}_b} w_{ab}(h) \label{eq:ns-conduction-sph-e}
\end{align}
where the source term $\dot{\rho}$ can be a "sink" of erupted vapor due to phase change like in ATHAM \citep{oberhuber1998volcanic}. $D$ is a drag force coefficient. $\kappa$ is heat conduction coefficient. Other physics can be added easily in a similar way. Adding of these new terms will lead to modification of several lines in the source code. The drag force term should show up only when dynamic dis-equilibrium between different phases is considered. In that case, each phase will need one set of governing equations of Navier-Stokes type. Adding of new phases into SPH code only needs few new lines for the new phase besides adding  interaction terms.

@@

\subsection{Time step}
The physical quantities (velocity, density, position and pressure) change every time step due to particle interactions. We have to consider the Courant condition, which is in spirit similar to the Courant condition for the grid-based methods, to determine the amount of the time step $\Delta t$.
\begin{equation}
\Delta t = CFL \min_a \bigg \lbrace \dfrac{[\frac{m_a}{\rho_a}]^{\frac{1}{d}}}{c_a} \bigg \rbrace
\end{equation}
Where $c_a$ is sound speed at particle a, $d$ is number of dimensions. First order Euler time integration, with $CFL = 0.2$ has been used in three dimensional calculations.

@@

Performance tests have been carried out on CCR (the computational cluster of Center for Computational Research) at Buffalo. E5645 CPUs running at 2.40GHz clock rate with 4GB memory per core on a Q-Logic Infiniband is used in these tests. Each node comprise of two sockets with 6 of these cores. Memory and level 3 cache are shared on each node. The initial domain is $[-4.8km,4.8km] \times [-4.8km,4.8km] \times [0km, 6km]$. Almost linear speed up is observed in our strong scalability test (Fig. \ref{fig:strong_scale}).
